---
title: "Playoffs 2 Data"
author: "Praveen Puviindran"
date: "2024-04-02"
output: pdf_document
---

```{r setup, include=FALSE}
library(readr)

nba_api <- read_csv("/Users/praveenpuviindran/Downloads/2019to2024nbagamestats.csv")

head(nba_api)
```

```{r}
# This script scrapes top-level NBA schedule and results from basketball-reference.com.
# User can set year and list of months to determine the window of games to scrape.
# At the end of the script, I reconstruct the conference standings based on W-L
# percentage.

library(rvest)
library(lubridate)
library(tidyverse)

########
# PARAMETERS
########
year <- "2024"
monthList <- c("october", "november", "december", "january", "february", 
               "march")
playoff_startDate <- ymd("2024-04-20")
outputfile <- "NBA-2024_game_data.rds"

########
# SCRIPT FOR SCRAPING DATA STARTS HERE
########
df <- data.frame()
for (month in monthList) {
  # get webpage
  url <- paste0("https://www.basketball-reference.com/leagues/NBA_", year, 
                "_games-", month, ".html")
  webpage <- read_html(url)
  
  # get column names
  col_names <- webpage %>% 
    html_nodes("table#schedule > thead > tr > th") %>% 
    html_attr("data-stat")    
  col_names <- c("game_id", col_names)
  
  # extract dates column
  # note that in april, there is a break in the table which just says 
  # "Playoffs". this messes with the data merging later, so we get rid of it
  dates <- webpage %>% 
    html_nodes("table#schedule > tbody > tr > th") %>% 
    html_text()
  dates <- dates[dates != "Playoffs"]
  
  # extract game id
  # we need to remove the NA that is due to the "Playoffs" row in april
  game_id <- webpage %>% 
    html_nodes("table#schedule > tbody > tr > th") %>%
    html_attr("csk")
  game_id <- game_id[!is.na(game_id)]
  
  # extract all columns (except date)
  data <- webpage %>% 
    html_nodes("table#schedule > tbody > tr > td") %>% 
    html_text() %>%
    matrix(ncol = length(col_names) - 2, byrow = TRUE)
  
  # combine game IDs, dates and columns in dataframe for this month, add col names
  month_df <- as.data.frame(cbind(game_id, dates, data), stringsAsFactors = FALSE)
  names(month_df) <- col_names
  
  # add to overall dataframe
  df <- rbind(df, month_df)
}

# change columns to the correct types
df$visitor_pts <- as.numeric(df$visitor_pts)
df$home_pts    <- as.numeric(df$home_pts)
df$attendance  <- as.numeric(gsub(",", "", df$attendance))
df$date_game   <- mdy(df$date_game)

# add column to indicate if regular season or playoff
df$game_type <- with(df, ifelse(date_game >= playoff_startDate, 
                                "Playoff", "Regular"))

# drop boxscore column
df$box_score_text <- NULL

# save to file
saveRDS(df, outputfile)

########
# SCRIPT FOR RANKING TABLE STARTS HERE
########

# get winner and loser of each game
df$winner <- with(df, ifelse(visitor_pts > home_pts, 
                             visitor_team_name, home_team_name))
df$loser <- with(df, ifelse(visitor_pts < home_pts, 
                            visitor_team_name, home_team_name))

# build up standings table for regular season
regular_df <- subset(df, game_type == "Regular")
regular_df <- na.omit(regular_df)
teams <- sort(unique(regular_df$visitor_team_name))
standings <- data.frame(team = teams, stringsAsFactors = FALSE)

# conference & division information: manually input
standings$conf <- c("East", "East", "East", "East", "East",
                    "East", "West", "West", "East", "West",
                    "West", "East", "West", "West", "West",
                    "East", "East", "West", "West", "East",
                    "West", "East", "East", "West", "West",
                    "West", "West", "East", "West", "East")
standings$div <- c("Southeast", "Atlantic", "Atlantic", "Southeast", "Central",
                   "Central", "Southwest", "Northwest", "Central", "Pacific",
                   "Southwest", "Central", "Pacific", "Pacific", "Southwest",
                   "Southeast", "Central", "Northwest", "Southwest", "Atlantic",
                   "Northwest", "Southeast", "Atlantic", "Pacific", "Northwest",
                   "Pacific", "Southwest", "Atlantic", "Northwest", "Southeast")

# populate W-L column, W pct
standings$win <- 0; standings$loss <- 0
for (i in 1:nrow(standings)) {
  standings$win[i]  <- sum(regular_df$winner == standings$team[i])
  standings$loss[i] <- sum(regular_df$loser  == standings$team[i])
}
standings$wl_pct <- with(standings, win / (win + loss))

# Eastern conference standings
east_standings <- subset(standings, conf == "East")
east_standings[with(east_standings, order(-wl_pct, team)), 
               c("team", "win", "loss")]

# Western conference standings
west_standings <- subset(standings, conf == "West")
west_standings[with(west_standings, order(-wl_pct, team)), 
               c("team", "win", "loss")]
```

```{r}
regular_df$home_prior_form <- nrow(regular_df)*NA

regular_df$visitor_prior_form <- nrow(regular_df)*NA



```

```{r}

for (i in 1:length(unique(regular_df$visitor_team_name))){
  

  adapted_df <- regular_df |> 
    filter(visitor_team_name == unique(regular_df$visitor_team_name)[i] | home_team_name == unique(regular_df$visitor_team_name)[i])
  

for (j in 1:(nrow(adapted_df) - 1)) 
  {

adapted_df$home_team_name[j+1] == unique(regular_df$visitor_team_name)[i]


if(adapted_df$visitor_team_name[j] == unique(regular_df$visitor_team_name)[i]){
#Are selected team away?
  
      if(is.na(adapted_df$visitor_prior_form[j])){
        #Is var column empty?
          if(adapted_df$winner[j] == unique(regular_df$visitor_team_name)[i]){
            if(adapted_df$visitor_team_name[j+1] == unique(regular_df$visitor_team_name)[i]){
            
            adapted_df$visitor_prior_form[j+1] <- "W"
            }
            else if (adapted_df$home_team_name[j+1] == unique(regular_df$visitor_team_name)[i]){
            adapted_df$home_prior_form[j+1] <- "W"

            }
        }
        if(adapted_df$loser[j]  == unique(regular_df$visitor_team_name)[i]){
          #Did selected team lose?
          if(adapted_df$visitor_team_name[j+1] == unique(regular_df$visitor_team_name)[i]){
            
            adapted_df$visitor_prior_form[j+1] <- "L"
            }
            else if (adapted_df$home_team_name[j+1] == unique(regular_df$visitor_team_name)[i]){
            adapted_df$home_prior_form[j+1] <- "L"

            }
        }
      }
      else if(nchar(adapted_df$visitor_prior_form[j]) < 9){
        
        if(adapted_df$winner[j] == unique(regular_df$visitor_team_name)[i]){
          if(adapted_df$visitor_team_name[j+1] == unique(regular_df$visitor_team_name)[i]){
          
          adapted_df$visitor_prior_form[j+1] <- paste(adapted_df$visitor_prior_form[j], "W")
       
          }
          else if (adapted_df$home_team_name[j+1] == unique(regular_df$visitor_team_name)[i]){
            adapted_df$visitor_prior_form[j+1] <- paste(adapted_df$visitor_prior_form[j], "W")
          }
        }
        else if(adapted_df$loser[j] == unique(regular_df$visitor_team_name)[i]){
          #Team lost
          if(adapted_df$visitor_team_name[j+1] == unique(regular_df$visitor_team_name)[i]){
          #Is team away next?
          adapted_df$visitor_prior_form[j+1] <- paste(adapted_df$visitor_prior_form[j], "L")
          
          }
          else if (adapted_df$home_team_name[j+1] == unique(regular_df$visitor_team_name)[i])
            #Is team at home next?
            {
            adapted_df$home_prior_form[j+1] <- paste(adapted_df$visitor_prior_form[j], "L")
          }
        }
        }
        
 else if (nchar(adapted_df$visitor_prior_form[j]) >= 9){
        if (adapted_df$winner[j] == unique(regular_df$visitor_team_name)[i]){
          #Did team win?
          
          if(adapted_df$visitor_team_name[j+1] == unique(regular_df$visitor_team_name)[i]){
            adapted_df$visitor_prior_form[j+1] <- paste(substr(adapted_df$visitor_prior_form[j], 3, 9), "W")
          }
          
          if(adapted_df$home_team_name[j+1] == unique(regular_df$visitor_team_name)[i]){
            adapted_df$home_prior_form[j+1] <- paste(substr(adapted_df$visitor_prior_form[j], 3, 9), "W")
          }
          
            
            }
        else if (adapted_df$loser[j] == unique(regular_df$visitor_team_name)[i]){
          
          
           if(adapted_df$visitor_team_name[j+1] == unique(regular_df$visitor_team_name)[i]){
            adapted_df$visitor_prior_form[j+1] <- paste(substr(adapted_df$visitor_prior_form[j], 3, 9), "L")
          }
          
          if(adapted_df$home_team_name[j+1] == unique(regular_df$visitor_team_name)[i]){
            adapted_df$home_prior_form[j+1] <- paste(substr(adapted_df$visitor_prior_form[j], 3, 9), "L")
          }
        }
      }
}





if(adapted_df$home_team_name[j] == unique(regular_df$visitor_team_name)[i]){
  
      if(is.na(adapted_df$home_prior_form[j])){
          if(adapted_df$winner[j] == unique(regular_df$visitor_team_name)[i]){
            if(adapted_df$visitor_team_name[j+1] == unique(regular_df$visitor_team_name)[i]){
            
            adapted_df$visitor_prior_form[j+1] <- "W"
            }
            else if (adapted_df$home_team_name[j+1] == unique(regular_df$visitor_team_name)[i]){
            adapted_df$home_prior_form[j+1] <- "W"

            }
          }
        if(adapted_df$loser[j]  == unique(regular_df$visitor_team_name)[i]){
          if(adapted_df$visitor_team_name[j+1] == unique(regular_df$visitor_team_name)[i]){
            
            adapted_df$visitor_prior_form[j+1] <- "L"
            }
            else if (adapted_df$home_team_name[j+1] == unique(regular_df$visitor_team_name)[i]){
            adapted_df$home_prior_form[j+1] <- "L"

            }
        }
      }
      else if(nchar(adapted_df$home_prior_form[j]) < 9){
        
        if(adapted_df$winner[j] == unique(regular_df$visitor_team_name)[i]){
          if(adapted_df$visitor_team_name[j+1] == unique(regular_df$visitor_team_name)[i]){
          
          adapted_df$visitor_prior_form[j+1] <- paste(adapted_df$home_prior_form[j], "W")
       
          }
          else if (adapted_df$home_team_name[j+1] == unique(regular_df$visitor_team_name)[i]){
            adapted_df$home_prior_form[j+1] <- paste(adapted_df$home_prior_form[j], "W")
          }
        }
        else if(adapted_df$loser[j] == unique(regular_df$visitor_team_name)[i]){
          if(adapted_df$visitor_team_name[j+1] == unique(regular_df$visitor_team_name)[i]){
          
          adapted_df$visitor_prior_form[j+1] <- paste(adapted_df$home_prior_form[j], "L")
       
          }
          else if (adapted_df$home_team_name[j+1] == unique(regular_df$visitor_team_name)[i]){
            adapted_df$home_prior_form[j+1] <- paste(adapted_df$home_prior_form[j], "L")
          }
        }
        }
        
 else if (nchar(adapted_df$home_prior_form[j]) >= 9){
        if (adapted_df$winner[j] == unique(regular_df$visitor_team_name)[i]){
          
          if(adapted_df$visitor_team_name[j+1] == unique(regular_df$visitor_team_name)[i]){
            adapted_df$visitor_prior_form[j+1] <- paste(substr(adapted_df$home_prior_form[j], 3, 9), "W")
          }
          
          if(adapted_df$home_team_name[j+1] == unique(regular_df$visitor_team_name)[i]){
            adapted_df$home_prior_form[j+1] <- paste(substr(adapted_df$home_prior_form[j], 3, 9), "W")
          }
          
            
            }
        else if (adapted_df$loser[j] == unique(regular_df$visitor_team_name)[i]){
          
          
           if(adapted_df$visitor_team_name[j+1] == unique(regular_df$visitor_team_name)[i]){
            adapted_df$visitor_prior_form[j+1] <- paste(substr(adapted_df$home_prior_form[j], 3, 9), "L")
          }
          
          if(adapted_df$home_team_name[j+1] == unique(regular_df$visitor_team_name)[i]){
            adapted_df$home_prior_form[j+1] <- paste(substr(adapted_df$home_prior_form[j], 3, 9), "L")
          }
        }
      }
    }

    
    regular_df <- regular_df |> 
      rbind(adapted_df[j,])
    
}
}
#Add condition: away next game vs. home next game and then add it to column
    

```


```{r}

final_df <- na.omit(regular_df)

head(final_df)
```


```{r}
str_replace_all(nba_api$TEAM_NAME, "LA Clippers", "Los Angeles Clippers")


home2023_24 <- nba_api |> 
  right_join(final_df, join_by(GAME_DATE == date_game, TEAM_NAME == home_team_name)) |> 
    rename(opposition = visitor_team_name) 


away2023_24 <- nba_api |> 
  right_join(final_df, join_by(GAME_DATE == date_game, TEAM_NAME == visitor_team_name)) |> 
  rename(opposition = home_team_name) |> 
  relocate(32, .before = 31)

names(home2023_24) == names(away2023_24)

combined <- rbind(home2023_24, away2023_24)
combined
```

```{r}
#Sample code to find monthly averages




combined |> 
  filter(format(GAME_DATE, "%m") == "03") |> 
  group_by(TEAM_NAME) 
```

```{r}
# Load necessary libraries
library(dplyr)
library(readr)

# Load the dataset
game_stats <- combined

# Inspect the dataset
head(game_stats)

```

```{r}

# Calculate Team Shooting Percentage
game_stats <- game_stats %>%
  mutate(Shooting_Percentage = FGM / FGA) %>%
  group_by(TEAM_NAME) %>%
  mutate(Avg_Shooting_Percentage = mean(Shooting_Percentage, na.rm = TRUE))

# Calculate Average OREB per Game
game_stats <- game_stats %>%
  group_by(TEAM_NAME) %>%
  mutate(Avg_OREB = mean(OREB, na.rm = TRUE))

# Opposition Team's Historic DREB (this would require a bit more complex logic to calculate historically for each game)
game_stats <- game_stats %>%
  group_by(MATCHUP) %>%
  mutate(Opponent_Avg_DREB = mean(DREB, na.rm = TRUE))

game_stats
```

```{r}
# Splitting the data into training and testing sets 
set.seed(123)
training_indices <- sample(seq_len(nrow(game_stats)), size = floor(0.8 * nrow(game_stats)))
training_data <- game_stats[training_indices, ]
testing_data <- game_stats[-training_indices, ]

# Build the model
model <- lm(OREB ~ Avg_Shooting_Percentage + Avg_OREB + Opponent_Avg_DREB, data = training_data)

# Summary of the model
summary(model)

# Predict and evaluate
predictions <- predict(model, newdata = testing_data)
mae <- mean(abs(predictions - testing_data$OREB))
print(paste("Mean Absolute Error:", mae))

```


```{r}
game_stats <- game_stats %>%
  mutate(
    # Shooting Efficiency: Points per Field Goal Attempt and Free Throw Attempt
    SE = PTS / (FGA + 0.44 * FTA), # The 0.44 factor adjusts for and-one's, technical foul shots, and three-shot fouls
    
    # Defensive Pressure: Combining Steals and Blocks
    DP = (STL + BLK) / MIN, # Normalized by minutes played
    
    # Home Advantage as a binary variable
    Home = ifelse(grepl("vs", MATCHUP), 1, 0)
  )

# Prepare for modeling: Remove any rows with NAs created during feature engineering
game_stats <- na.omit(game_stats)
game_stats$Pace = (game_stats$FGA + 0.44 * game_stats$FTA + game_stats$TOV + game_stats$OREB) / game_stats$MIN
# Estimate possessions based on available stats
game_stats$Possessions <- with(game_stats, 0.96 * (FGA + 0.44 * FTA - OREB + TOV))

# Calculate ORtg and DRtg
game_stats <- game_stats %>%
  mutate(ORtg = 100 * PTS / Possessions,
         DRtg = 100 * ifelse(Home == 1, visitor_pts, home_pts) / Possessions)

game_stats
```

```{r}
game_stats$HomeFormScore <- stringr::str_count(game_stats$home_prior_form, "W") - stringr::str_count(game_stats$home_prior_form, "L")
game_stats$VisitorFormScore <- stringr::str_count(game_stats$visitor_prior_form, "W") - stringr::str_count(game_stats$visitor_prior_form, "L")
game_stats$TS = game_stats$AST * game_stats$FG_PCT
#Team Synergy (TS): A metric to quantify the effectiveness of a team's collaboration, which might not be directly observable through traditional stats. It can be approximated by the interaction between assists and shooting percentage, assuming higher synergy leads to more efficient scoring opportunities.#
game_stats$FormDiff <- game_stats$HomeFormScore - game_stats$VisitorFormScore
game_stats <- game_stats %>%
  select(-game_start_time, -home_prior_form, -visitor_prior_form) %>%
  na.omit()
game_stats <- game_stats %>%
  mutate(Spread = ifelse(Home == 1, home_pts - visitor_pts, visitor_pts - home_pts),
         Total = home_pts + visitor_pts)
game_stats

```

```{r}
library(tidyverse)
library(caret)
library(lubridate)
library(randomForest)
library(neuralnet)
```

```{r}
set.seed(123)
model_spread_2 <- randomForest(Spread ~ ORtg + DRtg + Pace + DP + FormDiff, data = game_stats)
print(model_spread_2)
```
```{r}
team_stats_for_prediction <- data.frame(
  ORtg = 110,
  DRtg = 105,
  Pace = 95,
  DP = 0.2,
  FormDiff = 3
)
```

```{r}
predicted_spread <- predict(model_spread_2, newdata = team_stats_for_prediction)

print(predicted_spread)
```

```{r}
set.seed(123)
advanced_rf_model <- randomForest(Spread ~ ORtg + DRtg + Pace + DP + FormDiff + TS, data = game_stats, ntree = 500, mtry = 4, importance = TRUE)
print(advanced_rf_model)

```

Mean of Squared Residuals (MSR): The MSR value is 2.557157, which is relatively low. This metric represents the average of the squared differences between the observed actual outcomes and the predictions made by the model. A lower MSR indicates that the model's predictions are, on average, very close to the actual game spreads, suggesting high accuracy in the predictions.

Percentage of Variance Explained: The model explains 98.96% of the variance in the Spread, which is exceptionally high. This metric reflects the model's ability to capture the underlying patterns and relationships in the data that contribute to the outcome variable. A value this high suggests that the model is highly effective at predicting game spreads based on the selected predictors.

Interpretation
Predictive Power: The model's high percentage of variance explained, coupled with the low mean of squared residuals, indicates that it is extremely effective at predicting the spread of NBA games. This suggests that the combination of variables used, including both individual performance metrics like ORtg and DRtg, as well as team dynamics indicators like TS, provides a comprehensive set of predictors for understanding game outcomes in terms of spread.

Variable Selection: The choice to include variables such as ORtg, DRtg, and Pace aligns with conventional wisdom about factors that influence game outcomes. The inclusion of DP and FormDiff captures aspects of defensive intensity and recent performance trends, respectively, which are intuitively relevant to predicting spreads. TS as a measure of team synergy introduces a novel aspect into the predictive model, implying that how well a team plays together (beyond individual stats) is crucial for game outcomes.

Model Robustness: The configuration of the Random Forest, with a substantial number of trees and a thoughtful selection of variables at each split, likely contributes to its robustness and high explanatory power. The model benefits from Random Forest's inherent strengths, such as handling non-linear relationships and interactions between variables without explicit specification.

Conclusion
This Random Forest model demonstrates outstanding performance in predicting NBA game spreads, as evidenced by its near-perfect variance explanation and low mean squared residuals. The results suggest that the model, with its current configuration and variable selection, effectively captures the complexities and dynamics influencing NBA game outcomes.

```{r}
library(xgboost)
# Prepare data for xgboost
dtrain <- xgb.DMatrix(data = model.matrix(~ ORtg + DRtg + Pace + DP + FormDiff + TS - 1, data = game_stats), label = game_stats$Spread)
set.seed(123)
xgb_model <- xgb.train(data = dtrain, 
                       booster = "gbtree", 
                       objective = "reg:squarederror", 
                       nrounds = 1000, 
                       eta = 0.01,
                       max_depth = 6)
xgb.importance(model = xgb_model)


```
Interpretation of Each Feature's Importance:
Offensive Rating (ORtg):

Gain: With approximately 50.15% gain, ORtg is the most impactful feature in the model, significantly contributing to the model's predictive accuracy. This indicates that offensive efficiency is crucial for predicting game outcomes.
Cover: ORtg also has a high cover value, suggesting it influences a large portion of the observations in the dataset.
Frequency: Being used in 37% of splits, it's the most commonly selected feature for making decisions, highlighting its importance.
Defensive Rating (DRtg):

Gain: Nearly equal to ORtg in terms of gain, indicating that defensive efficiency is almost as critical as offensive efficiency in determining game outcomes.
Cover: Slightly lower than ORtg but still very influential, affecting a considerable number of observations.
Frequency: Used slightly less frequently than ORtg but still one of the top features for making splits.
Pace:

Gain: Minimal impact on model's predictive accuracy, indicating that the game's pace is not as crucial for predicting outcomes as the efficiency ratings.
Cover: Moderate cover value suggests it affects a fair number of observations when used.
Frequency: Despite its low gain, it's used relatively frequently, possibly indicating its role in interactions with other features or specific contexts.
Team Synergy (TS):

Gain: Lower than the efficiency ratings, implying it has a smaller but still meaningful contribution to model accuracy.
Cover and Frequency: Lower values than ORtg and DRtg, indicating it's less influential but still relevant.
Defensive Pressure (DP):

Gain: Very low, suggesting minimal contribution to improving model accuracy.
Cover and Frequency: Least among the features, reinforcing its minimal impact on the model.
FormDiff:

Gain: The lowest, indicating its contribution to model accuracy is negligible compared to other features.
Cover and Frequency: Lowest values, suggesting it's rarely influential in the model.
Overall Conclusion
The model strongly emphasizes the critical roles of Offensive and Defensive Ratings (ORtg and DRtg) in predicting game outcomes, with both features dominating in terms of gain, cover, and frequency. This underscores the balance between a team's offensive and defensive capabilities as the primary determinants of game performance in the NBA.

Pace, Team Synergy (TS), Defensive Pressure (DP), and Form Difference (FormDiff), while included, have significantly less impact on the model's predictions, with Pace being somewhat more frequently used despite its low gain. 

```{r}
game_stats$ShEff = (game_stats$FG_PCT + game_stats$FG3_PCT + game_stats$FT_PCT) / 3
#Shooting Efficiency (ShEff): A composite metric combining field goal, three-point, and free throw percentages to capture overall shooting effectiveness.
game_stats
```

```{r}
game_stats$FTImpact <- with(game_stats, FTA * FT_PCT)
#1. Free Throw Impact (FTImpact)- A measure combining free throw attempt rate and free throw percentage to assess the overall impact of free throws on the team's scoring.
game_stats$ShootingVersatilityIndex <- with(game_stats, 0.4 * FG_PCT + 0.4 * FG3_PCT + 0.2 * FT_PCT)
#2. Shooting Versatility Index (SVI)- This index aims to quantify a team's shooting versatility by combining field goal, three-point, and free throw percentages, giving higher weight to three-point efficiency.
game_stats$BallMovementIndex <- with(game_stats, AST / FGM)
#3. Ball Movement Index (BMI)- An index reflecting the team's ball movement, inferred from the ratio of assists to field goals made.
game_stats$DefensiveDisruptionIndex <- with(game_stats, (STL + BLK) / Possessions * 100)
#4. Defensive Disruption Index (DDI)- A composite measure reflecting the team's ability to disrupt the opponent's offense, calculated by combining steals and blocks relative to the number of possessions.
game_stats$ClutchFactor <- ifelse(abs(game_stats$PLUS_MINUS) <= 5 & game_stats$MIN >= 43, 1, 0)
#6. Clutch Factor (CF)- Designed to capture a team's performance in the clutch, defined as the last five minutes of a game with a point difference of 5 or less.
game_stats$PaceAdjustedDefImpact <- with(game_stats, DRtg / Pace)
#7. Pace Adjusted Defensive Impact (PADI)- Considers the defensive rating adjusted for the game's pace, offering insights into a team's defensive effectiveness relative to the game's speed.
game_stats
```
